---
title: "R Palooza with Kelly"
author: "Joscelin Rocha-Hidalgo"
date: "`r format(Sys.time(), '%a/%d/%b')`"
format:
  revealjs:
    self-contained: true
    slide-number: true
    theme: white
    transition: fade
    code-fold: false
editor: visual
---

```{r setup, include=FALSE}
library(gt)
library(palmerpenguins)
library(tidyverse)
library(gtExtras)

penguins_summary <- penguins %>%
  group_by(species) %>%
  summarize(
    mean_bill_length = round(mean(bill_length_mm, na.rm = TRUE),3),
    mean_bill_depth = round(mean(bill_depth_mm, na.rm = TRUE),3),
    mean_flipper_length = round(mean(flipper_length_mm, na.rm = TRUE),3)
  )
```

# [gt](https://gt.rstudio.com/)

![](https://gt.rstudio.com/reference/figures/gt_parts_of_a_table.svg)

## Creating a simple table

```{r}
#| echo: true
#| output-location: fragment
penguins_summary %>%
  gt() %>%
  tab_header(title = "Palmer Penguin Measurements")
```

## Renaming Columns

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "4-9"
penguins_summary %>%
  gt() %>%
  tab_header(title = "Palmer Penguin Measurements") %>%
  cols_label(
    species = "Species",
    mean_bill_length = "Bill Length (mean)",
    mean_bill_depth = "Bill Depth (mean)",
    mean_flipper_length = "Flipper Length (mean)"
  )
```

## Formatting Numbers

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "4-7"
penguins_summary %>%
  gt() %>%
  tab_header(title = "Palmer Penguin Measurements") %>%
  fmt_number(
    columns = c(mean_bill_length, mean_bill_depth, mean_flipper_length),
    decimals = 2
  ) %>%
  cols_label(
    species = "Species",
    mean_bill_length = "Bill Length (mean)",
    mean_bill_depth = "Bill Depth (mean)",
    mean_flipper_length = "Flipper Length (mean)"
  )
```

## Adding a Column Label

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "4-7"
penguins_summary %>%
  gt() %>%
  tab_header(title = "Palmer Penguin Measurements") %>%
  tab_spanner(
    label = "Mean Measurements (mm)",
    columns = c(mean_bill_length, mean_bill_depth, mean_flipper_length)
  ) %>%
  cols_label(
    species = "Species",
    mean_bill_length = "Bill Length",
    mean_bill_depth = "Bill Depth",
    mean_flipper_length = "Flipper Length"
  )
```

## Changing Alignment

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "13-17"
penguins_summary %>%
  gt() %>%
  tab_header(title = "Palmer Penguin Measurements") %>%
  tab_spanner(
    label = "Mean Measurements (mm)",
    columns = c(mean_bill_length, mean_bill_depth, mean_flipper_length)
  ) %>%
  cols_label(
    species = "Species",
    mean_bill_length = "Bill Length",
    mean_bill_depth = "Bill Depth",
    mean_flipper_length = "Flipper Length"
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  )
```

## Apply Color Themes to Table Cells

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-6"
penguins_summary %>%
  gt() %>%
  data_color(
    columns = c(mean_bill_length, mean_flipper_length),
    fn = scales::col_numeric(palette = "Blues", domain = NULL)
  )
```

## Customize Font Sizes and Styles

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-6"
penguins_summary %>%
  gt() %>%
  tab_options(
    table.font.size = px(14),
    table.font.names = "Arial"
  )
```

## Add Row Highlighting Based on Conditions

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-8"
penguins_summary %>%
  gt() %>%
  tab_style(
    style = cell_fill(color = "lightgray"),
    locations = cells_body(
      rows = mean_flipper_length > 200
    )
  )
```

## Add Borders for Columns or Rows

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-10"
penguins_summary %>%
  gt() %>%
  tab_style(
    style = cell_borders(
      sides = c("top", "bottom"),
      color = "black",
      weight = px(2)
    ),
    locations = cells_column_labels()
  )
```

## Another example

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-25"
penguins_summary %>%
  gt() %>%
  tab_style(
    style = cell_borders(
      sides = c("top", "bottom"),
      color = "black",
      weight = px(2)),
    locations = cells_column_labels())%>%
  tab_style(
    style = cell_borders(
      sides = c("bottom"),
      color = "black",
      weight = px(2)),
    locations = cells_body(rows = nrow(penguins_summary)))%>%
    tab_options(
    table.border.top.style = "none",
    table.border.bottom.style = "none",
    table_body.hlines.style = "none")
```

## Add Custom Row Labels

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "2-3"
penguins_summary %>%
  gt(rowname_col = "species") %>%
  tab_stubhead(label = "Penguin Species")

```

## Format Numbers with Patterns and Decimals

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-7"
penguins_summary %>%
  gt() %>%
  fmt_number(
    columns = c(mean_bill_length, mean_bill_depth, mean_flipper_length),
    decimals = 2,
    pattern = "{x} mm"
  )
```

## Add Group Labels and Spanners

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-6"
penguins_summary %>%
  gt() %>%
  tab_spanner(
    label = "Bill Measurements (mm)",
    columns = c(mean_bill_length, mean_bill_depth)
  ) %>%
    tab_spanner(
    label = "Flipper Measurements (mm)",
    columns = c(mean_flipper_length)
  ) %>%
    cols_label(
    species = "Species",
    mean_bill_length = "Length",
    mean_bill_depth = "Depth",
    mean_flipper_length = "Length"
  )
```

## Add Custom Footnotes

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-8"
penguins_summary %>%
  gt() %>%
  tab_footnote(
    footnote = "Data represents mean values per species.",
    locations = cells_column_labels(
      columns = c(mean_bill_length, mean_flipper_length)
    )
  )
```

## Add visualization to the table

```{r}
# Summarize mean bill length by species and year
penguin_summary <- penguins %>%
  drop_na(bill_length_mm, year) %>% # Ensure no missing data
  group_by(species, year) %>%
  summarize(mean_bill_length = mean(bill_length_mm), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = mean_bill_length, names_prefix = "Year_")

# Add a trend column with all raw bill length values across years for each species
penguin_trends <- penguin_summary %>%
  left_join(
    penguins %>%
      drop_na(bill_length_mm) %>%
      group_by(species) %>%
      summarize(trend = list(bill_length_mm), .groups = "drop"),
    by = "species"
  )
```

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "9-12"
penguin_trends %>%
  gt(rowname_col = "species") %>%
  cols_label(
    Year_2007 = "2007", # Adjust based on actual years in the dataset
    Year_2008 = "2008", # Adjust based on actual years in the dataset
    Year_2009 = "2009", # Adjust based on actual years in the dataset
    trend = "Density"
  ) %>%
  gt_plt_dist(
    column = trend, # Specify the column with the list of values
    type = "density" # Use density plots
  ) %>%
  tab_header(
    title = "Penguin Bill Length Trends",
    subtitle = "Mean bill lengths by year and density plot of all values"
  )
```

# [kable + kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)

## A Simple Table

```{r}
#| echo: true
#| output-location: fragment
library(kableExtra)
penguin_summary %>%
  kable() %>%
  kable_styling()
```

## Grouping Rows

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "5"
penguins_summary %>%
  group_by(species) %>%
  kable() %>%
  kable_styling() %>%
  add_header_above(c("Species Data" = 4))
```

## Adding a Caption

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "2"
penguins_summary %>%
  kable(caption = "Penguin Bill Measurements") %>%
  kable_styling()
```

## Adding Row Numbers

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "2"
penguins_summary %>%
  kable(row.names = TRUE) %>%
  kable_styling()
```

## Adding a Column Header Spanning Multiple Columns

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  kable() %>%
  add_header_above(c(" " = 1, "Bills" = 2, "Flippers"))
```

## Adding Borders to the Table

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-5"
penguins_summary %>%
  kable() %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(1, border_right = TRUE)
```

## Adding Bold and Italic Text

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-5"
penguins_summary %>%
  kable() %>%
  row_spec(1, bold = TRUE) %>%
  row_spec(3, italic = TRUE)
```

## Adding a Footer to the Table

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-5"
penguins_summary %>%
  kable() %>%
  kable_styling(font_size = 20) %>%
  add_footnote("Measurements are in millimeters.", , notation = "alphabet"
  )
```

## Adding Multiple Footnotes

```{r,results = "asis"}
# the asis and cat are needed for running it in quarto
#| echo: true
#| output-location: slide
#| code-line-numbers: "4-7"
penguins_summary %>%
  kable() %>%
  kable_styling(font_size = 35) %>%
  footnote(general = c("Measurements are in millimeters."),
           number = c("Note 1 with numbers.","Note 2 with numbers."),
           alphabet = c("Note 1 with letters.","Note 2 with letters."),
           symbol = c("Note 1 with symbols.","Note 2 with symbols.")) %>%
  cat
```

## Adding a Light Theme

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-5"
penguins_summary %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Adding Header with Style

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-5"
penguins_summary %>%
  kable() %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Measurements" = 3)) %>%
  row_spec(0, bold = TRUE, color = "white", background = "black")
```

## Highlighting Specific Rows

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  kable() %>%
  row_spec(2, bold = TRUE, color = "red")
```

## Coloring Columns

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  kable() %>%
  column_spec(2, color = "red")
```

## Customizing Font Size

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  kable() %>%
  kable_styling(font_size = 20)

```

# [**reactablefmtr**](https://kcuilla.github.io/reactablefmtr/index.html)

## A Simple Reactable Table

```{r}
#| echo: true
#| output-location: fragment
library(reactable)
library(reactablefmtr)

reactable(penguins_summary)
```

## Adding Column Labels

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-6"
reactable(
  penguins_summary,
  columns = list(
    mean_bill_length = colDef(name = "Mean Bill Length (mm)"),
    mean_bill_depth = colDef(name = "Mean Bill Depth (mm)"),
    mean_flipper_length = colDef(name = "Mean flipper Length (mm)")))
```

## Adding Conditional Formatting (Color Scales)

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "1,5-9"

color_set <- c("#f7c844","#429460","#2e6d9e")

reactable(
  penguins_summary,
  columns = list(
    mean_bill_length = colDef(
      cell = color_tiles(penguins_summary, 
                         color_set))))
```

## Highlighting Specific Rows

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-9"
reactable(
  penguins_summary,
  rowStyle = function(index) {
    if (penguins_summary$mean_bill_length[index] > 45) {
      list(background = "lightblue")
    } else {
      NULL
    }
  }
)
```

## Adding Row Names

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-4"
reactable(
  penguins_summary,
  rownames = TRUE
)
```

## Adding Borders to the Table

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-4"
reactable(
  penguins_summary,
  bordered = TRUE
)
```

## Adding Pagination

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-5"
reactable(
  penguins,
  pagination = TRUE,
  defaultPageSize = 3
)
```

## Adding Bars

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-4"
reactable(
  penguins_summary,
  columns = list(
    mean_flipper_length = colDef(
      maxWidth = 280,
      align = "center",
      cell = data_bars(
        data = penguins_summary,
        fill_color = c("#f7c844", "#429460", "#2e6d9e"),
        number_fmt = scales::number_format(accuracy = 0.2)
      ),
      style = list(borderLeft = "1px dashed rgba(0, 0, 0, 0.3)")
    )
  )
)
```

## Adding a Search Box

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-5"
reactable(
  penguins_summary,
  pagination = TRUE,
  defaultPageSize = 3,
  searchable = TRUE,
)
```

# [Flextable](https://davidgohel.github.io/flextable/index.html)

![](https://rfortherestofus.com/images/2019/11/flextable_functions.svg)

## A Simple Table

```{r}
#| echo: true
#| output-location: fragment
library(flextable)

penguins_summary %>%
  flextable()
```

## Adding a Header Title

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"

penguins_summary %>%
  flextable() %>%
  set_caption("Penguins Summary Table")
```

## Adding a Footer

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  add_footer_lines("This table summarizes penguin measurements.")
```

## Renaming Column Headers

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-8"
penguins_summary %>%
  flextable() %>%
    set_header_labels(
    mean_bill_length = "Mean Bill Length (mm)",
    mean_bill_depth = "Mean Bill Depth (mm)",
    mean_flipper_length = "Mean Flipper Length (mm)"
  )
```

## Highlighting Specific Rows

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3-8"
penguins_summary %>%
  flextable() %>%
    highlight(i = 1, color = "yellow")
```

## Adding Conditional Formatting

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
    color(i = ~ mean_bill_length < 48, color = "red")
```

## Adding Row Numbers

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "1,4"
penguins_summary$row_num <- seq_len(nrow(penguins_summary))

flextable(penguins_summary) %>%
  colformat_int(j = "row_num")
```

## Merging Cells

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "2"
penguins %>%
  slice(1:10) %>%
  flextable() %>%
  merge_v(j = c("island","year"))
```

## Changing Text Alignment

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
   align(j = "mean_bill_length", align = "center")
```

. . .

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
   align(j = c("mean_bill_length","row_num"), align = c("center","left"))
```

## Adding Background Colors

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
   bg(j = "species", bg = "lightblue")
```

## Customizing Text Colors

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  color(j = "mean_bill_depth", color = "orange")
```

## Adjusting Font Size

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  fontsize(part = "body", size = 20)
```

## Making Text Bold

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  bold(j = "species", bold = TRUE)
```

## Rotating Text

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
    rotate(j = "species", rotation = "btlr")
```

## Applying a Theme

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  theme_vanilla()
```

. . .

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  theme_box()
```

------------------------------------------------------------------------

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  theme_alafoli()
```

. . .

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  theme_vader()
```

. . .

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
  theme_apa()
```

## Adding Zebra Stripes

```{r}
#| echo: true
#| output-location: fragment
#| code-line-numbers: "3"
penguins_summary %>%
  flextable() %>%
    bg(i = seq(1, nrow(penguins_summary), by = 2), bg = "gray90")
```
